Computational design of RNA-based oscillatory circuits

### 2015-09-14 MScWriteup.tex
fiddling with figures

\newcounter{MYtempeqncnt}
\renewcommand{\thefootnote}{\roman{footnote}}

@@

\subsection{The sRNA regulatory system}
\label{The sRNA regulatory system}
In bacteria, one mechanism by which gene expression is naturally regulated is as follows \cite{Soper2010}: In order for a bacterial mRNA to be translated into a protein, the Ribosome must initially bind to the mRNA (Fig. \ref{RBS}). This occurs at the Ribosome Binding Site (RBS) \cite{Shine1974}, a specific nucleotide sequence found on the mRNA.

\begin{figure*}[t!]
\centering
\includegraphics[trim = 30 600 30 0,page=4,clip = true]{pnas1203831109.pdf}
\caption{A logical AND gate formed from a self repressed mRNA, and an sRNA which uncovers its RBS. In this system, transcription of the sRNA (transRNA) and mRNA (5'-UTR,GFP) are controlled by two promoter regions, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$. These are disabled by the presence of two chemical repressors, TetR and LacI, found naturally in the strain of \textit{E. coli} discussed. These chemical repressors are themselves disabled by two chemicals, aTC and IPTG. In the notation of the diagram, a barred line indicates repression, and an arrowed line indicates production. We see a `double negative' in aTc repressing TetR, which itself represses transcription of the sRNA (likewise for IPTG and the mRNA). Thus the presence of the sRNA and mRNA are controlled by the presence of aTc and IPTG, which can be experimentally introduced to the cell.  Image reproduced from \cite{Rodrigo2012}.}
\label{ANDGate}
\end{figure*}

\begin{figure}[H]
\centering
\includegraphics[trim = 100 170 100 400,page=10,clip = true,scale = 0.6]{Appendix.pdf}
\caption{A mechanism by which sRNA can regulate gene expression. Initially, the 5' UTR of the mRNA is folded over the RBS, forming a loop and blocking Ribosome binding. The sRNA binds to this loop in the mRNA, causing a conformational change which uncovers the RBS, and allows translation to occur. Image reproduced from \cite{Rodrigo2012}.}
\label{RBS}
\end{figure}

In an mRNA there is an untranslated region of nucleotides at the 5' end of the molecule (the UTR), upstream of the RBS \footnote{Both DNA and RNA are directional - the backbone of the molecule is not symmetric. This gives the molecule two distinct ends, denoted 5' and 3'. Translation can only occur in the 5' to 3' direction, hence the meaning of `upstream'}. Translation may be self repressed by this `tail' folding over and binding across the RBS, forming a  loop in the mRNA and preventing the Ribosome from binding (Fig. \ref{RBS}). This self repression may be released with an sRNA which binds to this looped tail - the new conformation of the sRNA:mRNA complex uncovers the RBS, allowing the Ribosome to bind. In summary, the presence of the sRNA positively regulates gene expresssion.

@@

\begin{figure}
    \centering
		\includegraphics[trim = 280 0 0  85 ,page=17 ,clip = true, scale=0.4]{WCPM_Warwick-talk45-220115.pdf}
        \caption{An overview of the experimental setup which allows single cell fluorescences to be recorded over time. Shown are the bacterial growth chambers (labelled `Microfabricated Microchemstat device'), the microscope imaging them, and the software constructing fluorescence time series. Image reproduced from \cite{Jaramillo}. }
            \label{ExperimentalSetup_overview}
\end{figure}
\begin{figure}
    \centering
	\includegraphics[trim = 0 0 0  90 ,page=18 ,clip = true, scale=0.35]{WCPM_Warwick-talk45-220115.pdf}
        \caption{A diagram of the bacterial growth chambers shown in Fig. \ref{ExperimentalSetup_overview}. The schematic shows the chambers themselves, and the medium inlet where aTc and IPTG are let in. Shown also are photographs of a row of chambers, and an individual chamber with bacteria growing in it. Image reproduced from \cite{Jaramillo}.}
    \label{ExperimentalSetup_chambers}
\end{figure}

@@

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t),
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. The least squares error between the model prediction and an experimental time series is defined as
\begin{align}
J(\boldsymbol{\theta}) = \sum_{i =1}^{n} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2,
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots n$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\begin{align}
\min_{\boldsymbol{\theta}} \sum_{i =1}^{n} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2.
\label{eq:min}
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function may be rugged and contain many local minima, which a local optimisation algorithm will may get stuck in. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular demonstrates that several Evolutionary Algorithms perform well on a test problem involving a set of ODEs modelling a biological system. We choose one of those recommended, the CMA-ES \cite{Hansen2006,Hansen2011}, which in addition. 
\todo[inline]{why use CMA-ES? Some refs}

@@

\begin{figure*}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
    \includegraphics[scale = 0.26,clip = true, trim = 90 0 70 0]{13_9_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{13\_9} dataset.}
                \label{InitialResults_13_9_best}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.26,clip = true, trim = 90 0 70 0]{14_7_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{14\_7} dataset.}
        \label{InitialResults_14_7_best}
    \end{subfigure}
        \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24, clip = true, trim = 140 0 125 0]{13_9_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{13\_9} dataset.}
        \label{InitialResults_13_9_hist}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 140 0 125 0]{14_7_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{14\_7} dataset. }
        \label{InitialResults_14_7_hist}
    \end{subfigure}
    \caption{Parameter estimates and model predictions for the \textit{13\_9} and \textit{14\_7} datasets. Note that in the model predictions, aTc forcing is shown - IPTG concentration is constant at a level which saturates the cell's response. The forcing curve's height is schematic - aTc concentration is switched between off, and on a level which saturates the cell's response.}
\label{InitialResults}
\end{figure*}

@@

 \begin{figure}[h]	
 	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.29,clip = true, trim = 82 0 70 0]{13_9_partial_data_13_9_pred}
        \caption{Model trained on \textit{13\_9}  60/60 data only, full prediction}
        \label{CrossValidation_1} 
    \end{subfigure}
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3,clip = true, trim = 87 0 70 0]{14_7_pred_13_9_data}
        \caption{\textit{14\_7} prediction \textit{13\_9} data}
        \label{CrossValidation_2} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3,clip = true, trim = 87 0 70 0]{13_9_pred_14_7_data}
        \caption{\textit{13\_9} prediction \textit{14\_7} data}
        \label{CrossValidation_3} 
    \end{subfigure}
    \caption{Cross validating data by taking parameter estimates from one dataset, and using them to predict another. \ref{CrossValidation_1} shows model predictions for the full \textit{13\_9} dataset, when only trained on the first forcing period data. \ref{CrossValidation_2}, \ref{CrossValidation_3} show model predictions for fitting to one dataset, and predicting the other.}
\label{CrossValidation}   
\end{figure}

@@

These problems may stem from structural inadequacies in the model (often termed \textit{structural identifiability}), in which two different parameter sets can give identical model predictions \cite{Jimenez-Hornero2008,Grewal1976 }. If this is the case, no amount of experimental data will allow us to estimate parameters, and we must consider reformulating the model\footnote{A simple example of a structurally non - identifiable model is $y = \beta_{1}\beta_{2}x$, where we are given data $(x,y)$, and asked to estimate parameters $\beta_{1}$ and $\beta_{2}$ - we can see that, in principle, only the product $\beta_{1}\beta_{2}$ can ever be estimated, a problem no amount of data can fix.}. Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in, parameter effects may be weak, or highly correlated, but in other regimes this is not true. In this case, parameter estimates may be improved by taking data in more varied experimental conditions, or attempting to increase the number of model outputs observed - in our case, observing more of the state vector than simply $z$.

@@

 \begin{align}
S_{ij} = \hat{\theta}_{j} \frac{\partial z}{ \partial \theta_{j}}\Bigr|_{t_{i}},
\end{align}
where $S_{ij}$ is the derivative of the observed fluorescence, $z$, with parameter $\theta_{j}$, evaluated at timepoint $t_{i}$. Each column of $S$ is a time series of sensitivity co-efficients, which describe how sensitive $z$ is to perturbations in the parameter associated with that column, and at what times it is most sensitive. $\hat{\theta}_{j}$ is the value of the parameter that the derivative is being evaluated at. It is included to set the scale that parameters may vary at, to ensure that apparently small sensitivity values do not result from a poor choice of units.

@@

\begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.23,clip = true, trim = 110 0 60 0]{Sensitivity_unscaled}
        \caption{Unscaled}
        \label{SensitivityMatrix_unscaled} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.23,clip = true, trim = 70 0 60 0]{Sensitivity_scaled}
        \caption{Scaled}
                \label{SensitivityMatrix_scaled} 
    \end{subfigure}
    \caption{Sensitivity coefficients $S_{ij}$ evaluated about a set of estimated parameters from the \textit{13\_9} dataset, for a single oscillation. \ref{SensitivityMatrix_unscaled} shows them unscaled, \ref{SensitivityMatrix_scaled} shows them scaled by the norm of each column of $S$.} 
\label{SensitivityMatrix}   
\end{figure}

@@

\begin{figure}[h]	
    \centering
        \includegraphics[scale = 0.25,clip = true, trim = 95 0 60 0]{model_output}
        \caption{Model output for all state variables over two oscillations, using the parameter values giving the lowest error on the \textit{13\_9} dataset, and normalised to lie on the same scale. Note $g$ is not explicitly shown, but is simply a rescaling of $z$ and as such would lie over it. }
        \label{model_output} 
\end{figure}

@@

These results suggest we may simplify our model, by replacing $\beta m +f_{s}\beta c$ in \eqref{eq:p} with the forcing function used in \eqref{eq:s}, and removing \eqref{eq:s} - \eqref{eq:c} entirely - in other words, moving the forcing term directly to the translation step. The simplified model is shown in \eqref{eq:p_simplified} - \eqref{eq:z_simplified}

@@

\begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.25 ,clip = true, trim = 100 0 60 400]{13_9_hist_simplified}
        \caption{Histograms of estimated parameter values, using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}.}
        \label{13_9_hist_simplified} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale =  0.25 ,clip = true, trim = 100 0 60 400]{13_9_scatterplot_simplified}
        \caption{Scatterplots of estimated parameters shown in \ref{13_9_hist_simplified}  against one another}
                \label{13_9_scatterplot_simplified} 
    \end{subfigure}
    \caption{Histograms and scatterplots of parameter estimates, found using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}.} 
\label{SimplifiedModelResults}   
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[scale = 0.23,clip = true, trim = 90 0 60 0]{Sensitivity_scaled_simplified}
\caption{Using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}, normalised sensitivity coefficients $S_{ij}$ evaluated about a set of estimated parameters from the \textit{13\_9} dataset, for a single oscillation.}
\label{Sensitivity_scaled_simplified}
\end{figure}

@@

 \begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.23 ,clip = true, trim = 85 0 0 0]{Likelihood_profile_rough}
        \caption{ Colormap scaled }
        \label{Likelihood_profile_rough} 
    \end{subfigure}
    	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.23 ,clip = true, trim = 85 0 0 0]{Likelihood_profile_fine}
        \caption{ Error function profile, $\Theta = 5400$}
        \label{Likelihood_profile_fine} 
    \end{subfigure}
    \caption{Error plotted as a function of $\mu$ and $F$, $\Theta = 5400$} 
\label{SimplifiedModelLikelihood}   
\end{figure}

@@

On the other hand, if a good description of the fluorescence data is what is really of interest, the simplified model may be a better starting point. $\mu$ and $\Theta$ are both parameters which can be easily estimated independent of our datasets - a realistic range of values could be used to bound the error function profile shown in Fig. \ref{SimplifiedModelLikelihood}. Further modelling work may focus on improving the modelling of translation, the rate limiting step - \cite{Hu2015}, for example, uses a more detailed model than our single step description of translation. Any additional parameters introduced will likely act on the same timescale as $\mu$, and stand a better chance of being estimable.

@@

We may also consist of changing our methodology for parameter estimation. Using a least squares minimisation approach has several problems, the first of which is local minima. Using the CMA-ES will give better results than a local minimisation algorithm, but one can never be sure that a global optimum has been found - for example, \cite{Algorithms2003} compares several algorithms and finds, in the test case considered, none locate the true global optimum.
A related problem is that of working with point estimates of parameters rather than distributions. The CMA-ES will only ever find a single optimum, giving a point estimate of parameter values, with no uncertainty information. To surmount this problem, we may rerun the algorithm many times, from many starting locations, and interpret the resulting spread of parameter values found as an uncertainty \cite{Hu2015}, as we have done in this report. However this methodology is rather ad hoc, and we have seen in Fig. \ref{SimplifiedModelLikelihood} that it can give a misleading impression of how tightly estimated parameters are.

@@

\begin{figure}[H]
    \begin{subfigure}[c]{0.49\textwidth}
        \centering
    \includegraphics[scale = 0.24, clip = true, trim = 120 0 70 0]{13_9_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{13\_9} dataset. }
        \label{InitialResults_13_9_correlation}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24, clip = true, trim = 80 0 70 0]{14_7_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{14\_7} dataset.}
        \label{InitialResults_14_7_correlation}
    \end{subfigure}

>>>>

\newcounter{MYtempeqncnt}
\renewcommand{\thefootnote}{\roman{footnote}}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\subsection{The sRNA regulatory system}
\label{The sRNA regulatory system}
In bacteria, one mechanism by which gene expression is naturally regulated is as follows \cite{Soper2010}: In order for a bacterial mRNA to be translated into a protein, the Ribosome must initially bind to the mRNA (Fig. \ref{RBS}). This occurs at the Ribosome Binding Site (RBS) \cite{Shine1974}, a specific nucleotide sequence found on the mRNA. 
\setcounter{figure}{1}
\begin{figure*}[t!]
\centering
\includegraphics[trim = 30 600 30 0,page=4,clip = true]{pnas1203831109.pdf}
\caption{A logical AND gate formed from a self repressed mRNA, and an sRNA which uncovers its RBS. In this system, transcription of the sRNA (transRNA) and mRNA (5'-UTR,GFP) are controlled by two promoter regions, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$. These are disabled by the presence of two chemical repressors, TetR and LacI, found naturally in the strain of \textit{E. coli} discussed. These chemical repressors are themselves disabled by two chemicals, aTC and IPTG. In the notation of the diagram, a barred line indicates repression, and an arrowed line indicates production. We see a `double negative' in aTc repressing TetR, which itself represses transcription of the sRNA (likewise for IPTG and the mRNA). Thus the presence of the sRNA and mRNA are controlled by the presence of aTc and IPTG, which can be experimentally introduced to the cell.  Image reproduced from \cite{Rodrigo2012}.}
\label{ANDGate}
\end{figure*}
\setcounter{figure}{0}

\begin{figure}[H]
\centering
\includegraphics[trim = 100 170 100 400,page=10,clip = true,scale = 0.6]{Appendix.pdf}
\caption{A mechanism by which sRNA can regulate gene expression. Initially, the 5' UTR of the mRNA is folded over the RBS, forming a loop and blocking Ribosome binding. The sRNA binds to this loop in the mRNA, causing a conformational change which uncovers the RBS, and allows translation to occur. Image reproduced from \cite{Rodrigo2012}.}
\label{RBS}
\end{figure}
\setcounter{figure}{2}

In an mRNA there is an untranslated region of nucleotides at the 5' end of the molecule (the 5' UTR), upstream of the RBS \footnote{Both DNA and RNA are directional - the backbone of the molecule is not symmetric. This gives the molecule two distinct ends, denoted 5' and 3'. Translation can only occur in the 5' to 3' direction, hence the meaning of `upstream'}. Translation may be self repressed by this `tail' folding over and binding across the RBS, forming a  loop in the mRNA and preventing the Ribosome from binding (Fig. \ref{RBS}). This self repression may be released with an sRNA which binds to this looped tail - the new conformation of the sRNA:mRNA complex uncovers the RBS, allowing the Ribosome to bind. In summary, the presence of the sRNA positively regulates gene expresssion.

@@

\begin{figure}
    \centering
		\includegraphics[trim = 280 0 0  85 ,page=17 ,clip = true, scale=0.45]{WCPM_Warwick-talk45-220115.pdf}
        \caption{An overview of the experimental setup which allows single cell fluorescences to be recorded over time. Shown are the bacterial growth chambers (labelled `Microfabricated Microchemstat device'), the microscope imaging them (the surrounding apparatus - lens, condenser, Dichroic mirror, CCD etc.), and the software constructing fluorescence time series. Image reproduced from \cite{Jaramillo}. }
            \label{ExperimentalSetup_overview}
\end{figure}
\begin{figure}
    \centering
	\includegraphics[trim = 0 0 0  90 ,page=18 ,clip = true, scale=0.35]{WCPM_Warwick-talk45-220115.pdf}
        \caption{A diagram of the bacterial growth chambers shown in Fig. \ref{ExperimentalSetup_overview}. The schematic shows the chambers themselves, and the medium inlet where aTc and IPTG are let in. Shown also are photographs of a row of chambers, and an individual chamber with bacteria growing in it. Image reproduced from \cite{Jaramillo}.}
    \label{ExperimentalSetup_chambers}
\end{figure}

@@

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t),
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. The least squares error between the model prediction and an experimental time series is defined as
\begin{align}
 \sum_{i =1}^{n} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2,
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots n$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\begin{align}
\argmin_{\boldsymbol{\theta}} \sum_{i =1}^{n} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2.
\label{eq:min}
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function may be rugged and contain many local minima, which a local optimisation algorithm may get stuck in. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular demonstrates that several Evolutionary Algorithms perform well on a test problem involving a set of ODEs modelling a biological system. We choose one of those recommended, the CMA-ES \cite{Hansen2006,Hansen2011}, which in addition. 
\todo[inline]{why use CMA-ES? Some refs}

@@

\begin{figure*}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
    \includegraphics[scale = 0.24,clip = true, trim = 115 0 60 0]{13_9_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{13\_9} dataset.}
                \label{InitialResults_13_9_best}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 100 0 70 0]{14_7_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{14\_7} dataset.}
        \label{InitialResults_14_7_best}
    \end{subfigure}
        \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24, clip = true, trim = 140 0 125 0]{13_9_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{13\_9} dataset.}
        \label{InitialResults_13_9_hist}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 140 0 125 0]{14_7_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{14\_7} dataset. }
        \label{InitialResults_14_7_hist}
    \end{subfigure}
    \caption{Parameter estimates and model predictions for the \textit{13\_9} and \textit{14\_7} datasets. Note that in the model predictions, aTc forcing is shown - IPTG concentration is constant at a level which saturates the cell's response. The forcing curve's height is schematic - aTc concentration is switched between off, and a level which saturates the cell's response.}
\label{InitialResults}
\end{figure*}

@@

 \begin{figure}[h!]	
 	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3,clip = true, trim = 70 0 70 0]{13_9_partial_data_13_9_pred}
        \caption{Model trained on \textit{13\_9}  60/60 data only, full prediction}
        \label{CrossValidation_1} 
    \end{subfigure}
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3,clip = true, trim = 70 0 70 0]{14_7_pred_13_9_data}
        \caption{\textit{14\_7} prediction \textit{13\_9} data}
        \label{CrossValidation_2} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.3,clip = true, trim = 70 0 70 0]{13_9_pred_14_7_data}
        \caption{\textit{13\_9} prediction \textit{14\_7} data}
        \label{CrossValidation_3} 
    \end{subfigure}
    \caption{Cross validating data by taking parameter estimates from one dataset, and using them to predict another. \ref{CrossValidation_1} shows model predictions for the full \textit{13\_9} dataset, when only trained on the first forcing period data. \ref{CrossValidation_2}, \ref{CrossValidation_3} show model predictions for fitting to one dataset, and predicting the other.}
\label{CrossValidation}   
\end{figure}

@@

These problems may stem from structural inadequacies in the model (often termed \textit{structural identifiability}), in which two different parameter sets can give identical model predictions \cite{Jimenez-Hornero2008,Grewal1976 }. If this is the case, no amount of experimental data will allow us to estimate parameters, and we must consider reformulating the model\footnote{A simple example of a structurally non - identifiable model is $y = \beta_{1}\beta_{2}x$, where we are given data $(x,y)$, and asked to estimate parameters $\beta_{1}$ and $\beta_{2}$ - we can see that, in principle, only the product $\beta_{1}\beta_{2}$ can ever be estimated, a problem no amount of data can fix.}. Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in (eg. the forcing function we use, the components of the model's state we can observe), parameter effects may be weak, or highly correlated, but in other regimes this is not true. In this case, parameter estimates may be improved by taking data in more varied experimental conditions, or attempting to increase the number of model outputs observed - in our case, observing more of the state vector than simply $z$.

@@

 \begin{align}
S_{ij} = \hat{\theta}_{j} \frac{\partial z}{ \partial \theta_{j}}\Bigr|_{t_{i}},
\label{eq:Sensitivity}
\end{align}
where $S_{ij}$ is the derivative of the observed fluorescence, $z$, with parameter $\theta_{j}$, evaluated at timepoint $t_{i}$. Each column of $S$ is a time series of sensitivity co-efficients, which describe how sensitive $z$ is to perturbations in the parameter associated with that column, and at what times it is most sensitive. $\hat{\theta}_{j}$ is the value of the parameter that the derivative is being evaluated at. It is included to set the scale that parameters may vary at, to ensure that apparently small sensitivity values do not result from a poor choice of units.

@@

\begin{figure}[h]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.23,clip = true, trim = 110 0 60 0]{Sensitivity_unscaled}
        \caption{Unscaled}
        \label{SensitivityMatrix_unscaled} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.23,clip = true, trim = 70 0 60 0]{Sensitivity_scaled}
        \caption{Scaled}
                \label{SensitivityMatrix_scaled} 
    \end{subfigure}
    \caption{Sensitivity coefficients $S_{ij}$ \eqref{eq:Sensitivity}, evaluated about a set of estimated parameters from the \textit{13\_9} dataset, for a single oscillation. \ref{SensitivityMatrix_unscaled} shows them unscaled, \ref{SensitivityMatrix_scaled} shows them scaled by the norm of each column of $S$.} 
\label{SensitivityMatrix}   
\end{figure}

@@

\begin{figure}[h]	
    \centering
        \includegraphics[scale = 0.23,clip = true, trim = 95 0 60 0]{model_output}
        \caption{Model output for all state variables over two oscillations, using the parameter values giving the lowest error on the \textit{13\_9} dataset, and normalised to lie on the same scale. Note $g$ is not explicitly shown, but is simply a rescaling of $z$ and as such would lie over it. }
        \label{model_output} 
\end{figure}

@@

These results suggest we may simplify our model, by replacing $\beta m +f_{s}\beta c$ in \eqref{eq:p} with the forcing function used in \eqref{eq:s}, and removing \eqref{eq:s} - \eqref{eq:c} entirely - in other words, moving the forcing term directly to the translation step. The simplified model is shown in \eqref{eq:p_simplified} - \eqref{eq:z_simplified}:

@@

\begin{figure}[h!]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24 ,clip = true, trim = 100 0 60 400]{13_9_hist_simplified}
        \caption{Histograms of estimated parameter values, using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}.}
        \label{13_9_hist_simplified} 
    \end{subfigure}
    \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale =  0.24 ,clip = true, trim = 100 0 60 400]{13_9_scatterplot_simplified}
        \caption{Scatterplots of estimated parameters shown in \ref{13_9_hist_simplified}  against one another}
                \label{13_9_scatterplot_simplified} 
    \end{subfigure}
    \caption{Histograms and scatterplots of parameter estimates, found using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}.} 
\label{SimplifiedModelResults}   
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[scale = 0.23,clip = true, trim = 90 0 60 0]{Sensitivity_scaled_simplified}
\caption{Using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}, normalised sensitivity coefficients $S_{ij}$ evaluated about a set of estimated parameters from the \textit{13\_9} dataset, for a single oscillation.}
\label{Sensitivity_scaled_simplified}
\end{figure}

@@

 \begin{figure}[t!]	
	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24 ,clip = true, trim = 55 0 0 0]{Likelihood_profile_rough}
        \caption{ A basin of low error values in $\mu - F$ space. Note the colour scaling maps all values above $20$ to the maximum colour.}
        \label{Likelihood_profile_rough} 
    \end{subfigure}
    	\begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24 ,clip = true, trim = 55 0 0 0]{Likelihood_profile_fine}
        \caption{Zoomed in on the bottom of the basin in Fig.\ref{Likelihood_profile_rough}. Note the colour scaling maps all values above $5.7$ to the maximum colour.}
        \label{Likelihood_profile_fine} 
    \end{subfigure}
    \caption{Error function plotted as a function of $\mu$ and $F$, $\Theta = 5400$. Note the many disconnected local minima in the bottom of a basin of low error function values.} 
\label{SimplifiedModelLikelihood}   
\end{figure}

@@

On the other hand, if a good description of the fluorescence data is what is really of interest, the simplified model may be a better starting point. On the experimental side, $\mu$ and $\Theta$ are both parameters which can be easily estimated independent of our datasets - a realistic range of values could be used to bound the error function profile shown in Fig. \ref{SimplifiedModelLikelihood}. Further modelling work may focus on improving the modelling of translation, the rate limiting step - \cite{Hu2015}, for example, uses a more detailed model than our single step description of translation. Any additional parameters introduced will likely act on the same timescale as $\mu$, and stand a better chance of being estimable.

@@

We may also consider changing our methodology for parameter estimation. Using a least squares minimisation approach has several problems, the first of which is local minima. Using the CMA-ES will give better results than a local minimisation algorithm, but one can never be sure that a global optimum has been found - for example, \cite{Algorithms2003} compares several algorithms and finds, in the test case considered, none locate the true global optimum.
A related problem is that of working with point estimates of parameters rather than distributions. The CMA-ES will only ever find a single optimum, giving a point estimate of parameter values, with no uncertainty information. To surmount this problem, we may rerun the algorithm many times, from many starting locations, and interpret the resulting spread of parameter values found as an uncertainty \cite{Hu2015}, as we have done in this report. However this methodology is rather ad hoc, and we have seen in Fig. \ref{SimplifiedModelLikelihood} that it can give a misleading impression of how tightly estimated parameters are.

@@

\begin{figure}[H]
    \begin{subfigure}[c]{0.49\textwidth}
        \centering
    \includegraphics[scale = 0.235, clip = true, trim = 120 0 65 0]{13_9_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{13\_9} dataset. }
        \label{InitialResults_13_9_correlation}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.235, clip = true, trim = 80 0 70 0]{14_7_heatmap.eps}
        \caption{Correlation matrix of estimated parameter sets, from the  \textit{14\_7} dataset.}
        \label{InitialResults_14_7_correlation}
    \end{subfigure}
